A Mechanised Library for Generic (Co)Programming

### 2024-02-22 paper/paper.tex
Updates to paper

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
Section~\ref{sec:TODO}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, catamorphisms take an
\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
\begin{minted}{Coq}
Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkCont sx (fun e => f (kx e)))
  end.
\end{minted}

It is easy to show that this function is a proper morphism, and that it is
unique. We can use the fact that catamorphisms are proper morphisms to encode
them as a higher-order morphism.
\begin{minted}{Coq}
Lemma cata_arr1 `{eA : equiv A} (g : Alg F A)
  : forall x y, x =e y -> cata_f g x =e cata_f g y.
Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...

Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

Terminal coalgebras are defined analogously to initial algebras:
\begin{minted}{Coq}
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
\end{minted}

Their equivalence relation, and anamorphisms, and universal properties are
easy to encode:
\begin{minted}{Coq}
  Definition ana `{eA : equiv A} : Coalg F A ~> A ~> GFix...

  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{minted}

With terminal coalgebras, we can encode another form of initial algebras. In
particular, the initial algebras based on terminal coalgebras are the carrier
types of terminal coalgebras, plus a proof that they are finite:

\begin{minted}{Coq}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{minted}

Using this predicate, we can define the notion of \emph{recursive coalgebras}
as those coalgebras that build finite trees. Recursive coalgebras are
implemented in Coq recursively on the proof that they produce finite trees.
Such proofs will be erased when extracting code to OCaml. This is standard
practice when building general recursive functions in Coq\footnote{\dcas{This
technique had a name in the community, but I cannot remember. Something
predicates}}.
\begin{minted}{Coq}
Notation rana_f__ h :=
  (fix f x H :=
    let hx := h x in
    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).

>>>>

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \mintinline{Coq}{LFixR}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, catamorphisms take an
\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
\begin{minted}{Coq}
Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkCont sx (fun e => f (kx e)))
  end.
\end{minted}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{minted}{Coq}
Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
\end{minted}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f
\o l_out}). The proof is by induction on the input to \mintinline{Coq}{f} and
\mintinline{Coq}{cata g}.
\begin{minted}{Coq}
Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

Terminal coalgebras are defined analogously to initial algebras:
\begin{minted}{Coq}
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
\end{minted}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode:
\begin{minted}{Coq}
  Definition ana `{eA : equiv A} : Coalg F A ~> A ~> GFix...

  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{minted}
This definition, however, prevents us from composing catamorphisms and
anamorphisms. To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{minted}{Coq}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{minted}
Using this predicate, we can define the notion of \emph{recursive coalgebras}
as those that build finite trees. Recursive coalgebras are implemented in Coq
recursively on the proof that they produce finite trees.  Such proofs will be
erased when extracting code to OCaml. This is standard practice when building
general recursive functions in Coq\footnote{\dcas{This technique had a name in
the community, but I cannot remember. Something predicates}}.
\begin{minted}{Coq}
Notation rana_f__ h :=
  (fix f x H :=
    let hx := h x in
    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).
